---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

## Set Working Directory
```{r}
setwd("C:/Users/Jane Chen/OneDrive/Documents/Data Science/Coursera Files/Course 05 Reproducible Research/week2")
```

## Loading and Preprocess Data
1. Load the data into object activity. 
```{r}
activity <- read.csv("activity.csv")
head(activity)
```
2. Preprocess Data
```{r}
class(activity$steps)
class(activity$date)
class(activity$interval)
```
as the class of the date column is factor, so I am changing it to date, using the as.Date function
```{r}
activity$date <- as.Date(activity$date)
class(activity$date)
```

## What is mean total number of steps taken per day? 
1. Calculate the total number of steps taken per day
```{r}
sumbydate <- tapply(activity$steps, activity$date, sum)
head(sumbydate)
```
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
```{r}
hist(sumbydate, xlab = "Number of Steps", ylab = "Frequency", main = "")
title(main = "Histogram of Total Number of Steps taken Each Day")
```

3. Calculate and report the mean and median of the total number of steps taken per day. 
```{r}
mean <- mean(sumbydate, na.rm = TRUE)
median <- median(sumbydate, na.rm = TRUE)
```
So, the mean total number of steps taken per day is `r mean` and the median is `r median`. 

## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis). 
```{r}
intave <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
t1 <- data.frame(interval = unique(activity$interval), steps = intave)
head(t1)
plot(t1, type = "l")
```

2. Whichc 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxint <- match(max(t1$steps), t1$steps)
t1[maxint, ]
```
The `r t1$interval[maxint]`th interval contains the maximum number of steps on average across all the days in the dataset. 

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data. 

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
numna <- sum(is.na(activity$steps))
numna
```
There are `r numna` missing values in the dataset. 

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc. 

Assuming number of steps taken in each 5-minute interval are similar across all the days, the missing values are replaced by the average number of steps taken in the 5-minute interval across all days. A function impute() is written to replace all the missing value with the average number of steps of each interval. 
```{r}
impute <- function(data = activity) {
    intave <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
    t1 <- data.frame(interval = unique(activity$interval), steps = intave)
    nd <- data.frame(steps = data$steps, date = data$date, interval = data$interval)
    for (i in 1:nrow(data))
        if(is.na(data$steps[i]) == TRUE) {
            nd$steps[i] <- intave[match(data$interval[i], t1$interval)]
        } else {
            nd$steps[i] <- data$steps[i]
        }
    nd
}
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in. 
```{r}
nd <- impute(activity)
head(nd)
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignmnet? What is the impact of imputing missing data on the estimates of the total daily number of steps? 
```{r}
newsumbydate <- tapply(nd$steps, nd$date, sum)
head(newsumbydate)
hist(newsumbydate, xlab = "Number of Steps", ylab = "Frequency", main = "")
title(main = "Histogram of Total Number of Steps taken Each Day")
newmean <- mean(newsumbydate, na.rm = TRUE)
newmedian <- median(newsumbydate, na.rm = TRUE)
```
The mean of total number of steps taken per day is now `r newmean` and the median is now `r newmedian`, comparing to the original mean `r mean` and median `r median`. The mean has not changed and the median has increased by about 1 step. 


## Are there differences in activity patterns between weekdays and weekends? 
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part. 

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day. 
```{r}
weekday <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
weekend <- c("Saturday", "Sunday")
nd$day <- weekdays(nd$date)
for (i in 1:nrow(nd)) {
    if(nd$day[i] %in% weekday)
        nd$weekday[i] <- "weekday"
    else if (nd$day[i] %in% weekend)
        nd$weekday[i] <- "weekend"
}
head(split(nd, nd$interval), 1)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data. 
```{r}
par(mfrow = c(2, 1), mar = c(3, 4, 2, 1))
weekdaynd <- nd[nd$weekday == "weekday", ]
weekendnd <- nd[nd$weekday == "weekend", ]
avewday <- data.frame(interval = unique(nd$interval), steps = tapply(weekdaynd$steps, weekdaynd$interval, mean, na.rm = TRUE))
avewend <- data.frame(interval = unique(nd$interval), steps = tapply(weekendnd$steps, weekendnd$interval, mean, na.rm = TRUE))
plot(avewday, type = "l", main = "Weekday", xlab = "", ylab = "Steps")
mtext("Interval", side = 1, line = 2)
plot(avewend, type = "l", main = "Weekend", xlab = "", ylab = "Steps")
mtext("Interval", side = 1, line = 2)
```